Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2021-04-13 ms/main.tex
Change \tilde to \dot

Not all adults become parents, and we denote the frequency of phenotype $A$ among parents by $\tilde{p}$.
Therefore, the frequency $\hat{p}$ of  phenotype $A$ among juveniles (after selection and vertical and oblique transmission) is
\begin{equation}\label{eq:horizontal}
\begin{aligned}
\hat{p} \;=
\tilde{p} [v + (1-v)p] + (1-\tilde{p}) [(1-v)p] \;= 
v \tilde{p} + (1-v) p \;.
\end{aligned}
\end{equation}
Individuals are assumed to interact according to a \emph{prisoner's dilemma}.
Specifically, individuals interact in pairs; a cooperator suffers a fitness cost $0<c<1$, and its partner gains a fitness benefit $b$, where we assume $c<b$. \autoref{fig:illustration}a shows the payoff matrix, i.e. the fitness of an individual with phenotype $\phi_1$ when interacting with a partner of phenotype $\phi_2$.

Social interactions occur randomly:
two juvenile individuals with phenotype $A$ interact with probability~$\hat{p}^2$, two juveniles with phenotype $B$ interact with probability $(1-\hat{p})^2$, and two juveniles with different phenotypes interact with probability $2\hat{p}(1-\hat{p})$. 
Horizontal cultural transmission occurs between pairs of individuals from the same generation. 
It occurs between socially interacting partners with probability~$\alpha$, or between a random pair with probability~$1-\alpha$ (see~\autoref{fig:illustration}b).
%The interaction-transmission association $\alpha$ is therefore the fraction of population that receives (horizontal transmission) from the social interaction partner, and $1-\alpha$ receives randomly.
However, horizontal transmission is not always successful, as one partner may reject the other's phenotype.
The probability of successful horizontal transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively (\autoref{table:interactions}, \autoref{fig:illustration}c).
Then, the frequency $p'$ of phenotype $A$ among adults in the next generation, after horizontal transmission, is 
\begin{equation}\label{eq:nextgen_adults}
\begin{aligned}
p' \;= \;
& \hat{p}^2 \big[\alpha + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& \hat{p}(1-\hat{p}) \big[\alpha(1-T_B) + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& (1-\hat{p})\hat{p} \big[\alpha T_A + (1-\alpha) \hat{p} T_A \big] \;+ 
(1-\hat{p})^2 \big[(1-\alpha) \hat{p} T_A \big] \\
=\; & \hat{p}^2(T_B-T_A) \;+\; \hat{p}(1+T_A-T_B) \;.
\end{aligned}
\end{equation}
The frequency of $A$ among parents (i.e. after selection) follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \tilde{p}' =
& \hat{p}^2 (1+b-c) \big[\alpha + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& \hat{p}(1-\hat{p}) (1-c) \big[\alpha(1-T_B) + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& (1-\hat{p})\hat{p} (1+b) \big[\alpha T_A + (1-\alpha) \hat{p} T_A \big] \;+\; 
(1-\hat{p})^2 \big[(1-\alpha) \hat{p} T_A \big] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \autoref{fig:illustration}a and \autoref{table:interactions}, and the population mean fitness is
%\begin{equation} \label{eq:mean_fitness}
$\bar{w} =  1 + \hat{p}(b-c)$.
%\end{equation}
%\autoref{eq:nextgen_parents} can be simplified to
%\begin{equation}\label{eq:nextgen_parents_simplified}
%\begin{aligned}
%\bar{w} \tilde{p}' =
%& \hat{p}^2 (1+b-c) \big[1-(1-\hat{p})(1-\alpha)T_B)\big] \;+ \\
%& \hat{p}(1-\hat{p}) (1-c) \big[\hat{p}(1-\alpha)T_B+1-T_B \big] \;+ \\
%& (1-\hat{p})\hat{p} (1+b) \big[\hat{p}(1-\alpha) + \alpha \big] T_A \;+ \\
%& (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
%\end{aligned}
%\end{equation}
%where $\hat{p}=v\tilde{p}+(1-v)p$.
Starting from \autoref{eq:horizontal} with $\hat{p}'=v\tilde{p}'+(1-v)p'$, we substitute $p'$ from \autoref{eq:nextgen_adults} and $\tilde{p}'$ from \autoref{eq:nextgen_parents} and obtain
\begin{equation} \label{eq:nextgen_juveniles}
\begin{aligned}
\hat{p}' =
& \frac{v}{\bar{w}}\Big[\hat{p}^2(1+b-c)\Big(1-(1-\hat{p})(1-\alpha)T_B)\Big)\Big] \;+ \\
& \frac{v}{\bar{w}}\Big[ \hat{p}(1-\hat{p})(1-c)\big( \hat{p}(1-\alpha)T_B + 1 - T_B \big) \Big] \;+ \\
& \frac{v}{\bar{w}}\Big[ \hat{p}(1-\hat{p})(1+b)\big(\hat{p}(1-\alpha) + \alpha \big) T_A \Big] \;+ \\
& \frac{v}{\bar{w}}(1-\hat{p})^2\hat{p}(1-\alpha)T_A \;+\;
(1-v)\hat{p}^2(T_B-T_A) \;+\;
(1-v)\hat{p}(1+T_A-T_B) \;.
\end{aligned}
\end{equation}
\autoref{table:vars_params} lists the model variables and parameters.

@@

In the absence of oblique transmission, $v=1$, the only equilibria are the fixation states, $\tilde{p}=0$ and $\tilde{p}=1$, and cooperation will evolve from any initial frequency (i.e., $\tilde{p}'>\tilde{p}$) if inequality~\ref{eq:vert_hori_alpha1} applies (\autoref{fig:equilibria_v1}).
This is similar to case of microbe-induced cooperation studied by \citet{lewin2017microbes}; therefore when $v=1$, this remark is equivalent to their eq.~1.

@@

%%% Table: variables and parameters
\bigskip
\begin{table}[h]
\centering
\caption{\textbf{Model variables and parameters.}
}
\begin{tabular}{lll}
\toprule
Symbol & Description & Values \\ \cmidrule(r){1-3}
$A$ & Cooperator phenotype & \\
$B$ & Defector phenotype & \\
$p$ & Frequency of phenotype $A$ among adults & $[0,1]$ \\
$\tilde p$ & Frequency of phenotype $A$ among parents & $[0,1]$ \\
$\hat p$ & Frequency of phenotype $A$ among juveniles & $[0,1]$ \\
$v$ & Vertical transmission rate & $[0,1]$ \\
$c$ & Cost of cooperation & $(0,1)$ \\
$b$ & Benefit of cooperation & $c<b$ \\
$\alpha$ & Probability of interaction-transmission association & $[0,1]$ \\
$T_A, T_B$ & Horizontal transmission rates of phenotype $A$ and $B$ & $(0,1)$ \\
\\ \bottomrule
\end{tabular}
\label{table:vars_params}
\end{table}
%\bigskip
%%%

@@

We assume here that $v=1$, i.e. no oblique transmission, and therefore $\hat{p}=\tilde{p}$.
Denote the frequencies of the pheno-genotypes $AM$, $BM$, $Am$, and $Bm$ by $\vec{\tilde{p}}= (\tilde{p}_1$, $\tilde{p}_2$, $\tilde{p}_3$, $\tilde{p}_4)$. 
The frequencies of the pheno-genotypes in the next generation are defined by the recursion system, 
\begin{equation} \label{eq:next_gen_p_1}
  \begin{aligned}
	% p1
  \bar{w}\tilde{p}_1' = 
  & \tilde{p}_1 x (1+b-c)(1 - (1-\alpha_1)(1-x)T_B) \;+ \\
  & \tilde{p}_1(1-x)(1-c)(1-\alpha_1T_B x - T_B(1-x)) \;+ \\
  & \tilde{p}_2 x (1+b)T_A(x + \alpha_1(1-x)) \;+ \\
  & \tilde{p}_2(1-x)x(1-\alpha_1)T_A \;,
\\
	% p2
  \bar{w}\tilde{p}_2' = 
  & \tilde{p}_1 x (1+b-c)(1-\alpha_1)(1-x)T_B \;+ \\
  & \tilde{p}_1(1-x)(1-c)(\alpha_1 T_B + (1-\alpha_1)(1-x)T_B) \;+ \\
  & \tilde{p}_2 x (1+b)(1-\alpha_1 T_A(1-x) - T_A x) \;+ \\
& \tilde{p}_2(1-x)(1 - (1-\alpha_1) x T_A) \;, 
\\
	% p3
  \bar{w}\tilde{p}_3' =
  & \tilde{p}_3 x (1+b-c)(1 - (1-\alpha_2)(1-x)T_B) \;+ \\
  & \tilde{p}_3(1-x)(1-c)(1-\alpha_2 T_B x - T_B(1-x)) \;+ \\
  & \tilde{p}_4 x (1+b)T_A(x + \alpha_2 (1-x)) \;+ \\
  & \tilde{p}_4(1-x) x (1-\alpha_2)T_A \;, 
\\
	% p4
  \bar{w}\tilde{p}_4' =
  & \tilde{p}_3 x (1+b-c)(1-\alpha_2)(1-x)T_B \;+ \\
  & \tilde{p}_3(1-x)(1-c)(\alpha_2 T_B + (1-\alpha_2)(1-x)T_B) \;+ \\
  & \tilde{p}_4 x (1+b)(1-\alpha_2T_A(1-x)-T_A x ) \;+ \\
  & \tilde{p}_4(1-x)(1 - (1-\alpha_2) x T_A) \;,
  \end{aligned}
\end{equation}
%which we can write as 
%\begin{equation}
%\bar{w} \vec{\tilde{p}} = F(\vec{\tilde{p}}),
%\end{equation}
where $x=\tilde{p}_1+\tilde{p}_3$ is the total frequency of the cooperative phenotype $A$, and $\bar{w} = 1 + (b-c) x$ is the population mean fitness.% and $\vec{\tilde{p}} = (\tilde{p}_1,\tilde{p}_2,\tilde{p}_3,\tilde{p}_4)$.

The equilibrium where only allele $M$ is present is  $\vec{\tilde{p}^*} = (\tilde{p}^*, 1-\tilde{p}^*, 0, 0)$, where
\begin{equation} \label{eq:p_tilde_star_alpha_1}
\tilde{p}^*=
\frac{c(1-T_B) - b \alpha_1 T_A - (T_A - T_B)}{b(1-\alpha_1)(T_A-T_B)} \;,
\end{equation}
setting $\alpha=\alpha_1$ and $v=1$ in \autoref{eq:general_equilibrium}.
When $v=1$, $\tilde p^*$ is a feasible polymorphism ($0 < \tilde{p}^* < 1$) if $T_A<T_B$ and $\gamma_2<c<\gamma_1$ (Result~\ref{result:vert_obli_hori}).

The local stability of $\vec{\tilde{p}^*}$ to the introduction of allele $m$ is determined by the linear approximation $\cl^*$ of the transformation in \autoref{eq:next_gen_p_1} near $\vec{\tilde{p}^*}$ (i.e., the Jacobian of the transformation at the equilibrium).
$\cl^*$ is known to have a block structure, with the diagonal blocks occupied by the matrices $\cl^*_{in}$ and $\cl^*_{ex}$ \citep{Liberman1986modifiers,Altenberg2017} .
The latter is the external stability matrix: the linear approximation to the transformation near $\vec{\tilde{p}^*}$ involving only the pheno-genotypes $Am$ and $Bm$, derived from \autoref{eq:next_gen_p_1}, with $\bar{w}^*=1+(b-c)\tilde{p}^*$ as the stable population mean fitness,
\begin{equation} \label{eq:external_stability_matrix}
\begin{aligned}
&\cl^*_{ex} = 
 \frac{1}{\bar{w}^*} \begin{bmatrix}
	 l_{1 1} &
	 l_{1 2} \\
	 l_{2 1} &
	 l_{2 2} 
\end{bmatrix} = 
\frac{1}{\bar{w}^*} \begin{bmatrix}
\frac{\partial\bar{w}\tilde{p}_3'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) &
\frac{\partial\bar{w}\tilde{p}_3'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) \\
\frac{\partial\bar{w}\tilde{p}_4'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) &
\frac{\partial\bar{w}\tilde{p}_4'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) 
\end{bmatrix} \;= \\
& \frac{1}{\bar{w}^*} \begin{bmatrix}
		(1+b \tilde{p}^* -c)(1-T_B(1-\tilde{p}^*)) + b \tilde{p}^* \alpha_2 T_B (1-\tilde{p}^*) & 
		(1+b \tilde{p}^*) T_A \tilde{p}^* + b \tilde{p}^* \alpha_2 T_A(1-\tilde{p}^*) \\
		(1+b \tilde{p}^* - c) T_B(1-\tilde{p}^*) - b \tilde{p}^* \alpha_2 T_B (1-\tilde{p}^*) &
		(1+b \tilde{p}^*) (1-T_A \tilde{p}^*) - b \tilde{p}^* \alpha_2 T_A (1-\tilde{p}^*) \\
  	\end{bmatrix} \;.
\end{aligned}
\end{equation}

Because we assume that $\vec{\tilde{p}^*}$ is internally stable (i.e. locally stable to small perturbations in the frequencies of $AM$ and $BM$), the stability of 
$\vec{\tilde{p}^*}$ is determined by the eigenvalues of the external stability matrix $\cl^*_{ex}$.
This is a positive matrix, and due to the Perron-Frobenius theorem, the leading eigenvalue of $\cl^*_{ex}$ is real and positive.
Thus, if the leading eigenvalue is less (greater) than one, then the equilibrium $\vec{\tilde{p}^*}$ is externally stable (unstable) and allele $m$ cannot (can) invade the population of allele $M$. 
The eigenvalues of $\cl^*_{ex}$ are the roots of the characteristic polynomial, $R(\lambda)$,
%\begin{equation}\label{char_poly}
%R(\lambda) = 
%\lambda^2 -\lambda \frac{(X+Q)}{\bar{w^*}} + \frac{XQ - YZ}{\bar{w^*}^2} \;,
%\end{equation}
%where $X$, $Y$, $Z$, and $Q$ are defined in \autoref{eq:external_stability_matrix}.
which is a quadratic with a positive leading coefficient. Therefore, $\lim_{\lambda \to \pm \infty} R(\lambda) = \infty$, and the leading eigenvalue is less than one (implying stability) if and only if $R(1)>0$ and $R'(1)>0$.
Thus, a sufficient condition for external instability of $\vec{\tilde{p}^*}$ is $R(1) < 0$.
%\begin{equation} \label{eq:R1_negative}
%\begin{aligned}
%R(1) < 0 \Leftrightarrow 
%1 - \frac{(X+Q)}{\bar{w^*}} + \frac{XQ - YZ}{\bar{w^*}^2} < 0 \Leftrightarrow 
%\bar{w^*}^2 - \bar{w^*} (X+Q) + XQ - YZ < 0 \;.
%\end{aligned}
%\end{equation}

>>>>

Not all adults become parents, and we denote the frequency of phenotype $A$ among parents by $\dot{p}$.
Therefore, the frequency $\hat{p}$ of  phenotype $A$ among juveniles (after selection and vertical and oblique transmission) is
\begin{equation}\label{eq:horizontal}
\begin{aligned}
\hat{p} \;=
\dot{p} [v + (1-v)p] + (1-\dot{p}) [(1-v)p] \;= 
v \dot{p} + (1-v) p \;.
\end{aligned}
\end{equation}
Individuals are assumed to interact according to a \emph{prisoner's dilemma}.
Specifically, individuals interact in pairs; a cooperator suffers a fitness cost $0<c<1$, and its partner gains a fitness benefit $b$, where we assume $c<b$. \autoref{fig:illustration}a shows the payoff matrix, i.e. the fitness of an individual with phenotype $\phi_1$ when interacting with a partner of phenotype $\phi_2$.

Social interactions occur randomly:
two juvenile individuals with phenotype $A$ interact with probability~$\hat{p}^2$, two juveniles with phenotype $B$ interact with probability $(1-\hat{p})^2$, and two juveniles with different phenotypes interact with probability $2\hat{p}(1-\hat{p})$. 
Horizontal cultural transmission occurs between pairs of individuals from the same generation. 
It occurs between socially interacting partners with probability~$\alpha$, or between a random pair with probability~$1-\alpha$ (see~\autoref{fig:illustration}b).
%The interaction-transmission association $\alpha$ is therefore the fraction of population that receives (horizontal transmission) from the social interaction partner, and $1-\alpha$ receives randomly.
However, horizontal transmission is not always successful, as one partner may reject the other's phenotype.
The probability of successful horizontal transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively (\autoref{table:interactions}, \autoref{fig:illustration}c).
Then, the frequency $p'$ of phenotype $A$ among adults in the next generation, after horizontal transmission, is 
\begin{equation}\label{eq:nextgen_adults}
\begin{aligned}
p' \;= \;
& \hat{p}^2 \big[\alpha + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& \hat{p}(1-\hat{p}) \big[\alpha(1-T_B) + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& (1-\hat{p})\hat{p} \big[\alpha T_A + (1-\alpha) \hat{p} T_A \big] \;+ 
(1-\hat{p})^2 \big[(1-\alpha) \hat{p} T_A \big] \\
=\; & \hat{p}^2(T_B-T_A) \;+\; \hat{p}(1+T_A-T_B) \;.
\end{aligned}
\end{equation}
The frequency of $A$ among parents (i.e. after selection) follows a similar dynamic, but also includes the effect of natural selection, and is therefore
\begin{equation}\label{eq:nextgen_parents}
\begin{aligned}
\bar{w} \dot{p}' =
& \hat{p}^2 (1+b-c) \big[\alpha + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& \hat{p}(1-\hat{p}) (1-c) \big[\alpha(1-T_B) + (1-\alpha)\big(\hat{p} + (1-\hat{p})(1-T_B)\big)\big] \;+ \\
& (1-\hat{p})\hat{p} (1+b) \big[\alpha T_A + (1-\alpha) \hat{p} T_A \big] \;+\; 
(1-\hat{p})^2 \big[(1-\alpha) \hat{p} T_A \big] \;,
\end{aligned}
\end{equation}
where fitness values are taken from \autoref{fig:illustration}a and \autoref{table:interactions}, and the population mean fitness is
%\begin{equation} \label{eq:mean_fitness}
$\bar{w} =  1 + \hat{p}(b-c)$.
%\end{equation}
%\autoref{eq:nextgen_parents} can be simplified to
%\begin{equation}\label{eq:nextgen_parents_simplified}
%\begin{aligned}
%\bar{w} \dot{p}' =
%& \hat{p}^2 (1+b-c) \big[1-(1-\hat{p})(1-\alpha)T_B)\big] \;+ \\
%& \hat{p}(1-\hat{p}) (1-c) \big[\hat{p}(1-\alpha)T_B+1-T_B \big] \;+ \\
%& (1-\hat{p})\hat{p} (1+b) \big[\hat{p}(1-\alpha) + \alpha \big] T_A \;+ \\
%& (1-\hat{p})^2 \hat{p} (1-\alpha) T_A \;.
%\end{aligned}
%\end{equation}
%where $\hat{p}=v\dot{p}+(1-v)p$.
Starting from \autoref{eq:horizontal} with $\hat{p}'=v\dot{p}'+(1-v)p'$, we substitute $p'$ from \autoref{eq:nextgen_adults} and $\dot{p}'$ from \autoref{eq:nextgen_parents} and obtain
\begin{equation} \label{eq:nextgen_juveniles}
\begin{aligned}
\hat{p}' =
& \frac{v}{\bar{w}}\Big[\hat{p}^2(1+b-c)\Big(1-(1-\hat{p})(1-\alpha)T_B)\Big)\Big] \;+ \\
& \frac{v}{\bar{w}}\Big[ \hat{p}(1-\hat{p})(1-c)\big( \hat{p}(1-\alpha)T_B + 1 - T_B \big) \Big] \;+ \\
& \frac{v}{\bar{w}}\Big[ \hat{p}(1-\hat{p})(1+b)\big(\hat{p}(1-\alpha) + \alpha \big) T_A \Big] \;+ \\
& \frac{v}{\bar{w}}(1-\hat{p})^2\hat{p}(1-\alpha)T_A \;+\;
(1-v)\hat{p}^2(T_B-T_A) \;+\;
(1-v)\hat{p}(1+T_A-T_B) \;.
\end{aligned}
\end{equation}
\autoref{table:vars_params} lists the model variables and parameters.

@@

In the absence of oblique transmission, $v=1$, the only equilibria are the fixation states, $\dot{p}=0$ and $\dot{p}=1$, and cooperation will evolve from any initial frequency (i.e., $\dot{p}'>\dot{p}$) if inequality~\ref{eq:vert_hori_alpha1} applies (\autoref{fig:equilibria_v1}).
This is similar to case of microbe-induced cooperation studied by \citet{lewin2017microbes}; therefore when $v=1$, this remark is equivalent to their eq.~1.

@@

%%% Table: variables and parameters
\bigskip
\begin{table}[h]
\centering
\caption{\textbf{Model variables and parameters.}
}
\begin{tabular}{lll}
\toprule
Symbol & Description & Values \\ \cmidrule(r){1-3}
$A$ & Cooperator phenotype & \\
$B$ & Defector phenotype & \\
$p$ & Frequency of phenotype $A$ among adults & $[0,1]$ \\
$\dot p$ & Frequency of phenotype $A$ among parents & $[0,1]$ \\
$\hat p$ & Frequency of phenotype $A$ among juveniles & $[0,1]$ \\
$v$ & Vertical transmission rate & $[0,1]$ \\
$c$ & Cost of cooperation & $(0,1)$ \\
$b$ & Benefit of cooperation & $c<b$ \\
$\alpha$ & Probability of interaction-transmission association & $[0,1]$ \\
$T_A, T_B$ & Horizontal transmission rates of phenotype $A$ and $B$ & $(0,1)$ \\
\\ \bottomrule
\end{tabular}
\label{table:vars_params}
\end{table}
%\bigskip
%%%

@@

We assume here that $v=1$, i.e. no oblique transmission, and therefore $\hat{p}=\dot{p}$.
Denote the frequencies of the pheno-genotypes $AM$, $BM$, $Am$, and $Bm$ by $\vec{\dot{p}}= (\dot{p}_1$, $\dot{p}_2$, $\dot{p}_3$, $\dot{p}_4)$. 
The frequencies of the pheno-genotypes in the next generation are defined by the recursion system, 
\begin{equation} \label{eq:next_gen_p_1}
  \begin{aligned}
	% p1
  \bar{w}\dot{p}_1' = 
  & \dot{p}_1 x (1+b-c)(1 - (1-\alpha_1)(1-x)T_B) \;+ \\
  & \dot{p}_1(1-x)(1-c)(1-\alpha_1T_B x - T_B(1-x)) \;+ \\
  & \dot{p}_2 x (1+b)T_A(x + \alpha_1(1-x)) \;+ \\
  & \dot{p}_2(1-x)x(1-\alpha_1)T_A \;,
\\
	% p2
  \bar{w}\dot{p}_2' = 
  & \dot{p}_1 x (1+b-c)(1-\alpha_1)(1-x)T_B \;+ \\
  & \dot{p}_1(1-x)(1-c)(\alpha_1 T_B + (1-\alpha_1)(1-x)T_B) \;+ \\
  & \dot{p}_2 x (1+b)(1-\alpha_1 T_A(1-x) - T_A x) \;+ \\
& \dot{p}_2(1-x)(1 - (1-\alpha_1) x T_A) \;, 
\\
	% p3
  \bar{w}\dot{p}_3' =
  & \dot{p}_3 x (1+b-c)(1 - (1-\alpha_2)(1-x)T_B) \;+ \\
  & \dot{p}_3(1-x)(1-c)(1-\alpha_2 T_B x - T_B(1-x)) \;+ \\
  & \dot{p}_4 x (1+b)T_A(x + \alpha_2 (1-x)) \;+ \\
  & \dot{p}_4(1-x) x (1-\alpha_2)T_A \;, 
\\
	% p4
  \bar{w}\dot{p}_4' =
  & \dot{p}_3 x (1+b-c)(1-\alpha_2)(1-x)T_B \;+ \\
  & \dot{p}_3(1-x)(1-c)(\alpha_2 T_B + (1-\alpha_2)(1-x)T_B) \;+ \\
  & \dot{p}_4 x (1+b)(1-\alpha_2T_A(1-x)-T_A x ) \;+ \\
  & \dot{p}_4(1-x)(1 - (1-\alpha_2) x T_A) \;,
  \end{aligned}
\end{equation}
%which we can write as 
%\begin{equation}
%\bar{w} \vec{\dot{p}} = F(\vec{\dot{p}}),
%\end{equation}
where $x=\dot{p}_1+\dot{p}_3$ is the total frequency of the cooperative phenotype $A$, and $\bar{w} = 1 + (b-c) x$ is the population mean fitness.% and $\vec{\dot{p}} = (\dot{p}_1,\dot{p}_2,\dot{p}_3,\dot{p}_4)$.

The equilibrium where only allele $M$ is present is  $\vec{\dot{p}^*} = (\dot{p}^*, 1-\dot{p}^*, 0, 0)$, where
\begin{equation} \label{eq:p_dot_star_alpha_1}
\dot{p}^*=
\frac{c(1-T_B) - b \alpha_1 T_A - (T_A - T_B)}{b(1-\alpha_1)(T_A-T_B)} \;,
\end{equation}
setting $\alpha=\alpha_1$ and $v=1$ in \autoref{eq:general_equilibrium}.
When $v=1$, $\dot p^*$ is a feasible polymorphism ($0 < \dot{p}^* < 1$) if $T_A<T_B$ and $\gamma_2<c<\gamma_1$ (Result~\ref{result:vert_obli_hori}).

The local stability of $\vec{\dot{p}^*}$ to the introduction of allele $m$ is determined by the linear approximation $\cl^*$ of the transformation in \autoref{eq:next_gen_p_1} near $\vec{\dot{p}^*}$ (i.e., the Jacobian of the transformation at the equilibrium).
$\cl^*$ is known to have a block structure, with the diagonal blocks occupied by the matrices $\cl^*_{in}$ and $\cl^*_{ex}$ \citep{Liberman1986modifiers,Altenberg2017} .
The latter is the external stability matrix: the linear approximation to the transformation near $\vec{\dot{p}^*}$ involving only the pheno-genotypes $Am$ and $Bm$, derived from \autoref{eq:next_gen_p_1}, with $\bar{w}^*=1+(b-c)\dot{p}^*$ as the stable population mean fitness,
\begin{equation} \label{eq:external_stability_matrix}
\begin{aligned}
&\cl^*_{ex} = 
 \frac{1}{\bar{w}^*} \begin{bmatrix}
	 l_{1 1} &
	 l_{1 2} \\
	 l_{2 1} &
	 l_{2 2} 
\end{bmatrix} = 
\frac{1}{\bar{w}^*} \begin{bmatrix}
\frac{\partial\bar{w}\dot{p}_3'}{\partial \dot{p}_3}(\vec{\dot{p}}^*) &
\frac{\partial\bar{w}\dot{p}_3'}{\partial \dot{p}_4}(\vec{\dot{p}}^*) \\
\frac{\partial\bar{w}\dot{p}_4'}{\partial \dot{p}_3}(\vec{\dot{p}}^*) &
\frac{\partial\bar{w}\dot{p}_4'}{\partial \dot{p}_4}(\vec{\dot{p}}^*) 
\end{bmatrix} \;= \\
& \frac{1}{\bar{w}^*} \begin{bmatrix}
		(1+b \dot{p}^* -c)(1-T_B(1-\dot{p}^*)) + b \dot{p}^* \alpha_2 T_B (1-\dot{p}^*) & 
		(1+b \dot{p}^*) T_A \dot{p}^* + b \dot{p}^* \alpha_2 T_A(1-\dot{p}^*) \\
		(1+b \dot{p}^* - c) T_B(1-\dot{p}^*) - b \dot{p}^* \alpha_2 T_B (1-\dot{p}^*) &
		(1+b \dot{p}^*) (1-T_A \dot{p}^*) - b \dot{p}^* \alpha_2 T_A (1-\dot{p}^*) \\
  	\end{bmatrix} \;.
\end{aligned}
\end{equation}

Because we assume that $\vec{\dot{p}^*}$ is internally stable (i.e. locally stable to small perturbations in the frequencies of $AM$ and $BM$), the stability of 
$\vec{\dot{p}^*}$ is determined by the eigenvalues of the external stability matrix $\cl^*_{ex}$.
This is a positive matrix, and due to the Perron-Frobenius theorem, the leading eigenvalue of $\cl^*_{ex}$ is real and positive.
Thus, if the leading eigenvalue is less (greater) than one, then the equilibrium $\vec{\dot{p}^*}$ is externally stable (unstable) and allele $m$ cannot (can) invade the population of allele $M$. 
The eigenvalues of $\cl^*_{ex}$ are the roots of the characteristic polynomial, $R(\lambda)$,
%\begin{equation}\label{char_poly}
%R(\lambda) = 
%\lambda^2 -\lambda \frac{(X+Q)}{\bar{w^*}} + \frac{XQ - YZ}{\bar{w^*}^2} \;,
%\end{equation}
%where $X$, $Y$, $Z$, and $Q$ are defined in \autoref{eq:external_stability_matrix}.
which is a quadratic with a positive leading coefficient. Therefore, $\lim_{\lambda \to \pm \infty} R(\lambda) = \infty$, and the leading eigenvalue is less than one (implying stability) if and only if $R(1)>0$ and $R'(1)>0$.
Thus, a sufficient condition for external instability of $\vec{\dot{p}^*}$ is $R(1) < 0$.
%\begin{equation} \label{eq:R1_negative}
%\begin{aligned}
%R(1) < 0 \Leftrightarrow 
%1 - \frac{(X+Q)}{\bar{w^*}} + \frac{XQ - YZ}{\bar{w^*}^2} < 0 \Leftrightarrow 
%\bar{w^*}^2 - \bar{w^*} (X+Q) + XQ - YZ < 0 \;.
%\end{aligned}
%\end{equation}

$R(\lambda)$ is defined as a determinant, $R(\lambda)=\det(\cl^*_{ex} - \lambda \ci)$, where $\ci$ is the 2-by-2 identity matrix. 
Since multiplication by a positive factor doesn't change the sign, and using the properties of the determinant,  we have
\begin{equation} \label{eq:signR1}
\begin{aligned}
\sign R(1) =
& \sign\det(\cl^*_{ex} - \ci) =
  \sign (\bar{w}^*)^2 \det(\cl^*_{ex} - \ci) = \\
& \sign \det(\bar{w}^*\cl^*_{ex} - \bar{w}^* \ci) =
  \sign \det\begin{bmatrix} l_{1 1} - \bar{w}^* & l_{1 2} \\ l_{2 1} & l_{2 2} - \bar{w}^* \end{bmatrix} \;,
\end{aligned}
\end{equation}
where $l_{i j}$ are defined in \autoref{eq:external_stability_matrix}.
Adding the second row in \autoref{eq:signR1} to the first row,  which does not change the determinant, and substituting $\bar{w}^*=1+(b-c)\dot{p}^*$, we get
\begin{equation} 
\begin{aligned}
\sign R(1) =
& \sign\det\begin{bmatrix}
	 -c(1-\dot{p}^*) &&
	 c \dot{p}^* \\
	 (1-\dot{p}^*)\big[(1+b\dot{p}^*-c)T_B - b \alpha_2 T_B \dot{p}^* \big] &&
	 \dot{p}^*\big[-(1+b\dot{p}^*)T_A - b \alpha_2 T_A (1-\dot{p}^*) + c\big]
\end{bmatrix} = \\
& \sign \Bigg[c \dot{p}^* (1-\dot{p}^*) \cdot \det\begin{bmatrix}
	 -1 &&
	 1 \\
	 (1+b\dot{p}^*-c)T_B - b \alpha_2 T_B \dot{p}^* &&
	 -(1+b\dot{p}^*)T_A - b \alpha_2 T_A (1-\dot{p}^*) + c
\end{bmatrix}\Bigg] = \\
& \sign\det\begin{bmatrix}
	 -1 &&
	 1 \\
	 (1+b\dot{p}^*-c)T_B - b \alpha_2 T_B \dot{p}^* &&
	 -(1+b\dot{p}^*)T_A - b \alpha_2 T_A (1-\dot{p}^*) + c
\end{bmatrix} \;,
\end{aligned}
\end{equation}
since $c>0$, $0<\dot{p}^*<1$.
That is, 
\begin{equation}
\begin{aligned}
\sign R(1) = 
&\sign\Big[
(1+b\dot{p}^*)T_A + b\alpha_2 T_A (1-\dot{p}^*) - c - 
(1+b\dot{p}^*-c)T_B + b\dot{p}^* \alpha_2 T_B \Big]
= \\
&\sign\Big[(1+b(1-\alpha_2)\dot{p}^*)(T_A-T_B) + b\alpha_2 T_A - c (1-T_B) \Big]\;.
\end{aligned}
\end{equation}
Substituting $\dot{p}^*$ from \autoref{eq:p_dot_star_alpha_1}, we get
\begin{equation} \label{eq:R1_less}
\begin{aligned}
R(1)<0 \Leftrightarrow
&\big[c(1-T_B) - b \alpha_1 T_A - (T_A - T_B)\big] \frac{1-\alpha_2}{1-\alpha_1} - c (1-T_B) + b\alpha_2 T_A  + (T_A-T_B) < 0 \Leftrightarrow \\
&(1-\alpha_2)\big[c(1-T_B) - b \alpha_1 T_A - (T_A - T_B)\big] < (1-\alpha_1)\big[c (1-T_B) - b\alpha_2 T_A  - (T_A-T_B) \big] \Leftrightarrow \\
& - b \alpha_1 T_A -\alpha_2 c(1-T_B) +\alpha_2(T_A - T_B) < 
  - b\alpha_2 T_A -\alpha_1 c (1-T_B) +\alpha_1 (T_A-T_B)  \Leftrightarrow \\
& \alpha_1 \big[c (1-T_B) - b T_A-(T_A-T_B)\big]  < 
  \alpha_2 \big[c(1-T_B)-b T_A -(T_A - T_B)\big] \Leftrightarrow \\
& \alpha_1 \big[b T_A + (T_A-T_B) - c (1-T_B)\big]  > 
  \alpha_2 \big[b T_A + (T_A-T_B) - c (1-T_B)\big] \;.
\end{aligned}
\end{equation}
We assumed $c<\gamma_1$, and since $0 \le \alpha_1 \le 1$,
\begin{equation} \label{eq:c_gamma1_utility}
\begin{aligned}
c &< \gamma_1 = \frac{b \alpha_1 T_A + (T_A-T_B)}{1-T_B} \; \Leftrightarrow \\
%c(1-T_B) < b \alpha_1 T_A + (T_A-T_B) \Leftrightarrow \\
0 &< b \alpha_1 T_A + (T_A-T_B) - c(1-T_B) \; \Rightarrow \\
0 &< b T_A + (T_A-T_B) - c(1-T_B) \;.
\end{aligned}
\end{equation}
